The XAI Paradox: Systems that Perform Well for the Wrong Reasons

### 2018-08-27 Paper.tex
Added example of datasets, experimental setup, noise discussion and the type of ML used

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. Furthermore, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

@@

The aim of this study is to investigate this question: to what extent do systems learn structures that make sense in a domain? In order to operationalize this idea, we generate structured datasets, where the embedded structure is a model of what makes sense in a domain. We investigate machine learning techniques in terms of how well they are able to learn the structure that defines datasets, rather than only in terms of a general performance accuracy. More specifically, artificial datasets will be generated from a set of predetermined formal conditions in terms of symmetric Boolean functions, on which machine learning systems will train. In this way, we can study how well the systems are able to extrapolate the correct conditions from the data. By combining multiple conditions defining datasets, we can investigate interaction effects whether the system learns each of the individual conditions correctly, or instead learns a different confounding structure that still accurately maps the input to the output.

@@

\subsection{Datasets}
%%%BV VERHELDEREN
For our experiment, a parameterized series of datasets is generated based on conditions defined in terms of symmetrical Boolean functions. Each instance in the dataset has an output value, which is true if and only if all of its conditions are true as well. This set of conditions can include the M-out-of-N function, the exact value function and the parity function. Each condition also has a set of parameters. As described earlier, parameter $n$ denotes the number of variables of the condition and parameter $m$ is used in M-out-of-N functions and exact value functions to determine the value of the output. The total number of variables of each instances in the datasets is therefore based on the number of variables of each of the conditions.

@@

\subsection{Machine Learning Algorithms}
Both a decision tree algorithm and a neural network are trained on the training sets. The resulting systems are tasked with classifying the general test sets and the specific test sets. The decision tree system uses the CART algorithm %%%BV REFERENCE INVOEGEN
with the Gini impurity and it does not make use of early stopping or any other form of pruning. The neural networks used consist of three hidden layers, with 25, 10 and 3 hidden nodes respectively. The learning rate is set to 0.05 and the sigmoid is used as the activation function. The networks use a mini-batch approach with a batch size of 50. The maximum number of times that the entire training dataset will be presented to the network during the training phase is set to 2000 rounds. Training stops early if there is no change in training errors over the past 10 rounds.

%%%BV INCLUDE A SUBSECTION `Experiments' HERE where you describe all experiments, including noise

@@

Both machine learning systems are first trained and tested using datasets with only a single condition; either the parity, M-out-of-N or exact value function. This will determine whether the systems are able to learn each of the conditions individually. The systems will then be trained and tested on datasets with two conditions in order to investigate both how well the systems are able to learn the combined problem and how well they are able to learn the individual conditions that make up the combined problem.

@@

Additionally, in order to investigate how well the systems perform with noise, a particular percentage of the instances of the training dataset are given a ``noisy variable''. These noisy variables have their values inverted, such that a 1 becomes a 0 and a 0 becomes a 1. It should be noted that inverting the value of a variable does not always mean that the outcome becomes incorrect. Varying the percentage of instances with a noisy variable between 0 and 100\% yields the graphs shown in Figure \ref{fig:SBF_noise}. These graphs display the mean accuracies of the decision tree (red) and the neural network (blue) on the specific test set for each of the three symmetrical Boolean functions.  The accuracies are average accuracies across values of $n$ ranging from 2 to 10 and across all possible values of $m$.

@@

Even under extreme levels of noise, the systems are still able to learn the M-out-of-N and exact value function quite effectively, showing the robustness of machine learning results. The parity function is a special case, as its output is dependent on whether the amount of 1's in the input vector is odd or even. This means that if an instance in the training dataset has a noisy variable, that instance automatically has the wrong output. If all of the instances in the training set have a noisy variable, the systems simply learn that an even number of 1's should yield true, which is the complete opposite of the parity function.%%%BV WHAT DO YOU WANT TO SAY HERE? CONNECTION TO DATA SHOWN?

>>>>

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. Furthermore, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

@@

The aim of this study is to investigate this question: to what extent do systems learn structures that make sense in a domain? In order to operationalize this idea, we generate structured datasets, where the embedded structure is a model of what makes sense in a domain. We investigate machine learning techniques in terms of how well they are able to learn the structure that defines datasets, rather than only in terms of a general performance accuracy. More specifically, artificial datasets will be generated from a set of predetermined formal conditions in terms of symmetric Boolean functions, on which neural networks and decision tree systems will train. In this way, we can study how well the systems are able to extrapolate the correct conditions from the data. By combining multiple conditions defining datasets, we can investigate interaction effects whether the system learns each of the individual conditions correctly, or instead learns a different confounding structure that still accurately maps the input to the output.

@@

\subsection{Datasets}
%%%BV VERHELDEREN
For our experiment, a parametrized series of datasets is generated based on conditions defined in terms of symmetrical Boolean functions. Each instance in the dataset has an output value, which is true if and only if all of its conditions are true as well. This set of conditions can include the M-out-of-N function, the exact value function and the parity function. Each condition also has a set of parameters. As described earlier, parameter $n$ denotes the number of variables of the condition and parameter $m$ is used in M-out-of-N functions and exact value functions to determine the value of the output. The total number of variables of each instances in the datasets is therefore based on the number of variables of each of the conditions. Table \ref{tbl:instance_example} shows two example instances of such a dataset. The instance of this dataset consist of two conditions: the parity condition with 2 variables (in red) and the M-out-of-N condition with 3 variables (in blue), where parameter $m$ is set to 2. In the first instance, the parity condition is false and the M-out-of-N condition is true. The output is therefore false, because all conditions  of an instance need to be true in order for the output to be true. In the second instance, the output is true, as both the parity condition (red) and the M-out-of-N condition (blue) are true.

\begin{table}[t]
\caption{Two example instances of a dataset with 2 parity variables (red) and 3 M-out-of-N variables (blue), where $m$=2.}
\centering
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\cellcolor[HTML]{FD6864}A & \cellcolor[HTML]{FD6864}B & \cellcolor[HTML]{34CDF9}C & \cellcolor[HTML]{34CDF9}D & \cellcolor[HTML]{34CDF9}E & Output \\ \hline
1                         & 1                         & 0                         & 1                         & 1                         & 0      \\ \hline
\end{tabular}
\centering
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\cellcolor[HTML]{FD6864}A & \cellcolor[HTML]{FD6864}B & \cellcolor[HTML]{34CDF9}C & \cellcolor[HTML]{34CDF9}D & \cellcolor[HTML]{34CDF9}E & Output \\ \hline
1                         & 0                         & 0                         & 1                         & 1                         & 1      \\ \hline
\end{tabular}
\label{tbl:instance_example}
\end{table}

@@

\subsection{Machine Learning Algorithms}
Both a decision tree algorithm and a neural network are trained on the training sets. The resulting systems are tasked with classifying the general test sets and the specific test sets. The decision tree system uses the CART algorithm \cite{breiman1984classification} with the Gini impurity and it does not make use of early stopping or any other form of pruning. The neural networks used consist of three hidden layers, with 25, 10 and 3 hidden nodes respectively. The learning rate is set to 0.05 and the sigmoid is used as the activation function. The networks use a mini-batch approach with a batch size of 50. The maximum number of times that the entire training dataset will be presented to the network during the training phase is set to 2000 rounds. Training stops early if there is no change in training errors over the past 10 rounds.

%%%BV INCLUDE A SUBSECTION `Experiments' HERE where you describe all experiments, including noise
\subsection{Experiments}
Both machine learning systems are first trained and tested using datasets with only a single condition; either the parity, M-out-of-N or exact value function. This will determine whether the systems are actually able to learn each of the conditions individually. The systems will then be trained and tested on datasets with two conditions, such as the example instances in Table \ref{tbl:instance_example}, in order to investigate both how well the systems are able to learn the combined problem and how well they are able to learn the individual conditions that make up the combined problem.

To show the ability of the systems perform under noise, a additional tests will be performed. In these experiments, a particular percentage of the instances of the training datasets are given a ``noisy variable''. These noisy variables have their values inverted, such that a 1 becomes a 0 and a 0 becomes a 1. After training on the datasets with noise, the systems are tasked with classifying the same test sets as before.

@@

@@

Varying the percentage of instances with a noisy variable between 0 and 100\% yields the graphs shown in Figure \ref{fig:SBF_noise}. These graphs display the mean accuracies of the decision tree (red) and the neural network (blue) on the specific test set for each of the three symmetrical Boolean functions.  The accuracies are average accuracies across values of $n$ ranging from 2 to 10 and across all possible values of $m$. It should be noted that inverting the value of a variable does not always mean that the outcome of the instance becomes incorrect. Therefore if 50\% of the instances have a noisy variable, it does not necessarily mean that 50\% of the instances have an incorrect output.

@@

Even under extreme levels of noise, the systems are still able to learn the M-out-of-N and exact value function quite effectively, showing the robustness of machine learning results.

%%%CS Leave this out entirely, or move it to discussion?
\iffalse
The M-out-of-N function is most robust to noise, as seen in Figure \ref{fig:SBF_MU}. This is because inverting one variable of the input vector only changes the output in two cases. First of all, if the total number of 1's is equal to $m$ and the output is true, changing a 1 to a 0 due to noise should change the output to false. Secondly, if the total number of 1's is equal to $m-1$ and the output is false, and the noise changes a variable from 0 to 1, the output should be true.

The accuracies on the exact function drop faster with more noise, as seen in Figure \ref{fig:SBF_EU}. When an instance of this dataset has a true output, and one of its variables is inverted, the total number of 1's will no longer be equal to $m$, and therefore the output will be incorrect. If an instance in the exact dataset has an output that is false, however, the noise can do two things: the number of 1's might be equal to $m$, thus creating a true output, or (the more likely alternative) the number of 1's may still not be equal to $m$, and the output is still false. This effect can be seen in Figure \ref{fig:SBF_EU}, as the systems are only able to correctly classify the incorrect instances, thus achieving an accuracy of roughly 50\%.

The parity function is a different case, as its output is dependent on whether the amount of 1's in the input vector is odd or even. By adding a noisy variable to an instance of the parity function, the amount of 1's will always change from odd to even, or even to odd. This means that every instance of a pairty dataset that has a noisy variable automatically has the wrong output. This is why the accuracies on the parity function drops more significantly than on the other functions (to 0\% in the case of the decision tree in Figure \ref{fig:SBF_PU}). 
\fi

%%%BV WHAT DO YOU WANT TO SAY HERE? CONNECTION TO DATA SHOWN?
